Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-01 analysis/paper/paper.Rmd
added length-weight section to code

```{r deadwater-attr}
# set some deadwater attributes
ddw_rkms = 1.871
ddw_m2 = 119431
ddw_avg_width_m = 68
ddw_acres = ddw_m2 / 4047
ddw_hectares = ddw_m2 / 10000

@@

```{r capture-data, eval = T}
raw_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # break days into "sampling events"
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

@@

```{r effort, eval = T}
# What size (mm) is considered a large pikeminnow?
large_size = 380

@@

```{r lincoln-petersen-estimator, eval = T}
# using Lincoln-Petersen model, combining first 4 and last four days
# define mark and recapture periods
cap_periods = tibble(event_name = "Fall_2019",
                     cap_period = "Mark",
                     int = interval(start = ymd('20191112'),
                                    end = ymd('20191114'))) %>%
  bind_rows(tibble(event_name = "Fall_2019",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20191119'),
                                  end = ymd('20191121')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Mark",
                   int = interval(start = ymd('20201020'),
                                  end = ymd('20201023')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20201027'),
                                  end = ymd('20201030'))))

@@

```{r assump-violation}
# assume that marked fish aren't available for recapture until 2 days after capture
# using Schnabel estimator
N_mods_viol = N_mods %>%
  filter(model == "Schnabel") %>%
  select(event_name:model) %>%
  mutate(data = map(data,
                    .f = function(x) {
                      x %>%
                        mutate(R_lag = lag(R),
                               R_lag = replace_na(R_lag,
                                                  0))
                    })) %>%
  mutate(mr_model = map2(data,
                         model,
                         .f = function(x, y) {
                           with(x,
                                mrClosed(n = n,
                                         m = m,
                                         R = R_lag,
                                         method = y,
                                         chapman.mod = TRUE))
                         }),
         N = map_dbl(mr_model,
                     .f = summary),
         CI = map(mr_model,
                  .f = confint),
         Lci = map_dbl(CI,
                       .f = function(x) x[1]),
         Uci = map_dbl(CI,
                       .f = function(x) x[2])) %>%
  ungroup() %>%
  select(-CI) %>%
  mutate(model = "Schnabel - Delayed Mixing")
```

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  bind_rows(N_mods_viol %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer"),
         model = fct_relevel(model,
                             "Petersen",
                             after = 0)) %>%
  arrange(event_name, model) %>%
  filter(! model %in% c("Petersen",
                        "Schumacher-Eschmeyer"))

@@

@@

# average size of fall and spring pikeminnow
avg_fall_NPM   = 352.9
avg_spring_NPM = 393.7

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995.

@@

This resulted in average spring and fall starting weights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 rkm downstream of Deadwater Slough.

@@

<!--
We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given a mean Northern Pikeminnow population size of `r prettyNum(unique(impact_df$N[impact_df$season == "Fall"]), big.mark = ",")` from fall estimates and `r prettyNum(unique(impact_df$N[impact_df$season == "Spring"]), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e., 50% of their diet consists of fish), we'd estimate that the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 90]), big.mark = ",")` g) fish during the DSR emigration and an additional `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 90]), big.mark = ",")` g) during the NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]) + unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g of fish during the DSR and NRR emigrations to maintain their population size if we assume 50% of their diet consists of fish.
-->

>>>>

```{r deadwater-attr}
# set some deadwater attributes
ddw_rkms = 1.871               # length km
ddw_m2 = 119431                # area square m
ddw_avg_width_m = 68           # measured avg width
ddw_acres = ddw_m2 / 4047
ddw_hectares = ddw_m2 / 10000

@@

```{r capture-data}
raw_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # break days into "sampling events"
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

@@

```{r effort}
# What size (mm) is considered a large pikeminnow?
large_size = 380

@@

```{r lincoln-petersen-estimator}
# using Lincoln-Petersen model, combining first 4 and last four days
# define mark and recapture periods
cap_periods = tibble(event_name = "Fall_2019",
                     cap_period = "Mark",
                     int = interval(start = ymd('20191112'),
                                    end = ymd('20191114'))) %>%
  bind_rows(tibble(event_name = "Fall_2019",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20191119'),
                                  end = ymd('20191121')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Mark",
                   int = interval(start = ymd('20201020'),
                                  end = ymd('20201023')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20201027'),
                                  end = ymd('20201030'))))

@@

```{r assump-violation}
# assume that marked fish aren't available for recapture until 48 days after release
# using Schnabel estimator
N_mods_viol = N_mods %>%
  filter(model == "Schnabel") %>%
  select(event_name:model) %>%
  mutate(data = map(data,
                    .f = function(x) {
                      x %>%
                        mutate(R_lag = lag(R),
                               R_lag = replace_na(R_lag,
                                                  0))
                    })) %>%
  mutate(mr_model = map2(data,
                         model,
                         .f = function(x, y) {
                           with(x,
                                mrClosed(n = n,
                                         m = m,
                                         R = R_lag,
                                         method = y,
                                         chapman.mod = TRUE))
                         }),
         N = map_dbl(mr_model,
                     .f = summary),
         CI = map(mr_model,
                  .f = confint),
         Lci = map_dbl(CI,
                       .f = function(x) x[1]),
         Uci = map_dbl(CI,
                       .f = function(x) x[2])) %>%
  ungroup() %>%
  select(-CI) %>%
  mutate(model = "Schnabel - Delayed Mixing")
```

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  bind_rows(N_mods_viol %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer"),
         model = fct_relevel(model,
                             "Petersen",
                             after = 0)) %>%
  arrange(event_name, model) %>%
  # remove the Petersen and Schumacher-Eschmeyer estimators
  filter(! model %in% c("Petersen",
                        "Schumacher-Eschmeyer"))

@@

```{r length-weight}
# average size of fall and spring pikeminnow
avg_fall_NPM_TL = cmr_df %>%
  filter(grepl('Fall', event_name)) %>%
  select(length) %>%
  pull() %>%
  mean(na.rm = T)

avg_spring_NPM_TL = cmr_df %>%
  filter(grepl('Spring', event_name)) %>%
  select(length) %>%
  pull() %>%
  mean(na.rm = T)

# conversions from Fishbase.org
#TL = 1.052 * FL
#TL = 1.087 * FL (12-30cm)
#TL = 1.114 * FL (5-12cm) don't use

# use mean of coefficients
# TL = 1.0695 * FL or
# FL = 0.9350 * TL

# convert to FL
avg_fall_NPM_FL = avg_fall_NPM_TL * 0.9350
avg_spring_NPM_FL = avg_spring_NPM_TL * 0.9350

# convert FL to g using Parker (1995) equation
avg_fall_NPM_g = 10^(-4.886 + (2.986*log10(avg_fall_NPM_FL)))
avg_spring_NPM_g = 10^(-4.886 + (2.986*log10(avg_spring_NPM_FL)))

@@

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average TL of Northern Pikeminnow caught in Deadwater Slough during our study was converted to fork length (FL) and then to weight (grams) using a weight-length formula from @Parker1995.

@@

This resulted in average fall and spring starting weights of `r round(avg_fall_NPM_g, 1)` and `r round(avg_spring_NPM_g, 1)` g for Northern Pikeminnow with average TLs of `r round(avg_fall_NPM_TL, 1)` and `r round(avg_spring_NPM_TL, 1)` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 rkm downstream of Deadwater Slough.

@@

<!--
We estimated an individual, average-size (`r avg_fall_NPM_TL` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM_TL` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given a mean Northern Pikeminnow population size of `r prettyNum(unique(impact_df$N[impact_df$season == "Fall"]), big.mark = ",")` from fall estimates and `r prettyNum(unique(impact_df$N[impact_df$season == "Spring"]), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e., 50% of their diet consists of fish), we'd estimate that the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 90]), big.mark = ",")` g) fish during the DSR emigration and an additional `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 90]), big.mark = ",")` g) during the NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]) + unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g of fish during the DSR and NRR emigrations to maintain their population size if we assume 50% of their diet consists of fish.
-->
